Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-04-12 bbbq_article.tex
Updates from Overleaf
Use shorthand names for haplotypes

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from human transmembrane helices (TMHs) 
are predicted to be presented more often than expected,
based on their abundance in the proteome.
We show that this over-presentation of TMH-derived peptides is general, 
as it is predicted for diverse bacteria and viruses 
and for both MHC-I and MHC-II.  
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show that TMHs are evolutionarily more conserved, 
as single nucleotide polymorphisms (SNPs) 
are present relatively less frequent in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more effective recognition of pathogens 
than the innate immune response, 
are T cells which recognize peptides derived from 
exogenous antigens presented on Major Histocompatibility Complexes (MHC) class I and II.

@@

The MHC proteins are heterodimeric proteins encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide-binding groove of MHC-I is made by the alpha chain, 
and there are three forms of alpha chains called HLA-A, HLA-B and HLA-C. 
For MHC-II, both the alpha and the beta chains contribute to the binding groove, 
and there are three major forms as well, called HLA-DR, HLA-DQ and HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA allele is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

While humans only express a limited number of MHC haplotypes, 
and therefore any individual's immune system detects 
only a fraction of all possible peptide fragments, 
the number of pathogenic peptides that can be detected 
is increased at the population level, 
because of the multiple and highly polymorphic MHC genes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development of prediction algorithms 
that allow for very reliable \emph{in silico} predictions 
of the binding affinities of peptides
(\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).
For example, \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to MHC,
86\% were confirmed to do so in vivo.

@@

Using these prediction algorithms, 
we recently predicted that peptides derived 
from transmembrane helices (TMHs) 
will be more frequently presented by MHC-I 
than expected based on their abundance (\cite{bianchi2017}). 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs. 
This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity 
to bind than hydrophilic ones.

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}),
for example, \cite{jones2007improving} found that,
from 184 TMPs with known topology, 
80\% of the TMH predictions replicated this finding.

TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH (\cite{ahram2006estimation}).
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMHs 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

This study has two objectives. 
First, we aimed to generalize our findings by predicting the presentation of peptides 
from different kingdoms of life and for both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, 
and for most haplotypes of both MHC-I and II. 
We confirmed this over-presentation of TMH-derived peptides 
by re-analysis of a peptide elution study. 
Second, we tested our hypothesis that TMHs 
are more evolutionarily conserved than solvent-exposed protein regions. 
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell branch of the adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

To predict how frequently epitopes overlapping with TMHs are presented,
we applied a similar analysis strategy as described in \cite{bianchi2017} 
for several haplotypes of both MHC-I and MHC-II, 
as well as for a human, viral and bacterial proteome.
To summarize, for each proteome, 
all possible 9-mers (for MHC-I) or 14-mers (MHC-II) were derived. 
For each of these peptides, we determined if it overlapped with a predicted 
TMH and if it was predicted to bind to each haplotype.

For MHC-I, 9-mers were used, as this is the length most frequently presented in MHC-I 
and was used in our earlier study (\cite{bianchi2017}). 
For MHC-II, 14-mers were used, 
as these are the most frequently occurring epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis} (MTb), UniProt ID UP000001584) 
reference proteome. 
We used TMHMM (\cite{krogh2001predicting}) to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; (\cite{shao2020high}) for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
For MHC-II, we selected haplotypes with a phenotypic frequency 
of at least 14\% in the human population (\cite{greenbaum2011functional}),
resulting in 21 MHC-II haplotypes.

The MHC-I molecule presents epitopes of shorter length than MHC-II.
We expect this to be of no influence in the presentation level
of TMH-derived epitopes, for each of the three proteomes.
We verified that MHC-II is just as prone to over-present TMH-derived
epitopes by comparing the relative over-presentation of these epitopes.

In our previous work, we found that the observed pattern can be
explained from the hydrophobicity of the MHC-I binding cleft (\cite{bianchi2017}).
Here we do a similar analysis, by correlating the percentage of TMH-derived
epitopes presented versus the mean hydrophobicity of all presented
peptides. From these points, we derive a linear trendline and measure
the proportion of the correlation explained by the assumed linear relation.

This study differs in one important aspect from our previous work (\cite{bianchi2017}), 
The definition of a binder differs from \cite{bianchi2017}:
in the current study a peptide is called a binder if, within a certain haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a \emph{proteome} 
(see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}, 
which put less strain on the bioinformatics pipeline.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
We believe that our revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify our results are similar,
we compare the previous and the
current approach side by side in the supplementary materials.

@@

To determine if epitopes derived from TMHs are presented \emph{in vivo}
we used two elution studies, both of which have sequenced 
the epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640\_9606.
For the epitopes that were present in the proteome exectly once,
we predicted the topology of the proteins in which these epitopes are found.
We used both TMHMM (\cite{krogh2001predicting}) and PureseqTM (\cite{wang2019efficient}) to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane region.

@@

To detect the evolutionary conservation of TMHs,
we collected human single nucleotide polymorphisms (SNPs)
that resulted in a single amino acid substitution,
and determined if this substitution occurred within a predicted TMH or not.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: the 'dbSNP' (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases 'gene' (for gene names, \cite{brown2015gene})
and 'protein' (for proteins sequences, \cite{sayers2010database}).

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) 
for the organism \emph{Homo sapiens}.
This resulted in 1,077 gene IDs (on December 2020).
% 1077 gene IDs is correct for December 2020.
% At 2021-03-01, one will get 1130 gene IDs.
% Also, one of the gene IDs that was valid back then,
% has been obsoleted.
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database was queried for 
variations associated with the gene name. 
As the NCBI API constrains its users to three calls per second
(to assure fair use), we had to limit the extent of our analysis.

@@

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positions of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.

Per SNP, the 'protein' NCBI database was queried for the
protein sequence.
For each protein sequence, the protein topology was determined 
using PureseqTM.
Using these predicted protein topologies, 
the SNPs were scored to be located within or outside trans membrane domains.

@@

Figure \ref{fig:bbbq_1_smart_results} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I haplotype, it shows the percentage of binders that overlap with a TMH 
for at least one residue.
A horizontal line shows the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as soluble regions.
In 11 out of 13 MHC-I haplotypes, TMH-derived epitopes are presented more often 
then the null expectation, for a human and bacterial proteome.
For the viral proteome, 12 out of 13 haplotypes present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the various different haplotypes 
is similar for the probed proteomes, 
which strengthens our previous conclusion (\cite{bianchi2017}) 
that the hydrophobicity of the MHC-binding groove 
is the main factor responsible for the predicted over-presentation 
of TMH-derived peptides.

@@

  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc1.png}
    \label{fig:hydrophobicity_1}
  \end{subfigure}  
  \hfill
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc2.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \caption{
    \textbf{(A)} 
    TMH-derived peptides are predicted to be over-presented in general,
    which is for MHC-I (left column) and MHC-II (right column),
    as well as for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The bars show the percentage of epitopes predicted to be 
    overlapping with TMHs.
    The red lines indicate the percentage as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \textbf{(B)} and \textbf{(C)}:
    The relation between the percentage of TMH-derived epitopes presented,
    and the hydrophobicity score of binders for MHC-I haplotypes (panel B)
    and MHC-II haplotypes (panel C). The trendline assumes a linear relationship.
  }
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II haplotypes, 
in a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{MTb}
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the exact TMH and epitope counts.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Hydrophocity explains the generality of this finding}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Hydrophobicity may be the common reason of the consistent over-presentation
of  (hydrophobic) TMH-derived epitopes.
Figures \label{fig:hydrophobicity_1} and \label{fig:hydrophobicity_2}
show the hydrophobicity preference score for the extent over-presentation
of TMH-derived epitopes for the different haplotypes in a human proteome.
An assumed linear correlation explains 88\% of the variability in MHC-I.
For MHC-II, 62\% of the variability is explained by hydrophobicity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{TMH-derived peptides are predicted to be presented in vivo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{bbbq_article_issue_157/results.png}
  \caption{
    TMH-derived epitopes are presented \emph{in vivo},
    for both MHC-I and MHC-II, for two topology prediction
    programs.
  }
  \label{fig:elution}
\end{figure}

To obtain experimental confirmation that peptides stemming from TMHs 
are presented in MHC-I and MHC-II,
two peptide elution studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} for MHC-I and MHC-II respectively) 
were reanalyzed.
Figure \ref{fig:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I and MHC-II elution 
studies,
for the two topology prediction tools TMHMM (\cite{krogh2001predicting}) and PureseqTM (\cite{wang2019efficient}). 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.
See the supplementary materials (table \ref{tab:elution}) for the exact values.

@@

From this study, two important conclusions can be drawn. 
First, the MHC over-presentation of TMHs is likely a general feature 
and predicted to occur for most haplotypes of both MHC-I and -II 
and for humans as well as bacterial and viral pathogens. 
Second, TMHs are more evolutionary conserved than solvent-exposed protein motifs, 
at least in the human proteome.

>>>>

Cytolytic T cell responses are predicted to be biased towards membrane proteins. The peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic,  therefor peptide fragments derived from human transmembrane helices (TMHs) are predicted to be presented more often as would be expected 
based on their abundance in the proteome. However, the physiological reason of why membrane proteins might be over-presented is unclear. 
In this study, we show that the over-presentation of TMH-derived peptides is general, as it is predicted for bacteria and viruses 
and for both MHC-I and MHC-II. Moreover, we show that TMHs are evolutionarily more conserved, because single nucleotide polymorphisms (SNPs) are present relatively less frequently in TMH-coding chromosomal regions compared to regions coding for extracellular and cytoplasmic protein regions. Thus, our findings suggest that both cytolytic and helper T cells respond more to membrane proteins, because these are evolutionary more conserved. We speculate that TMHs therefor are less prone to escape mutations that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more specific recognition of pathogens 
than the innate immune response, 
are T cells which recognize peptides, called epitopes, derived from antigenic proteins presented on Major Histocompatibility Complexes (MHC) class I and II on the cell surface.

@@

The MHC proteins are heterodimeric complexes encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide binding groove of MHC-I is made by only the alpha subunit. There are three classical forms of MHC-I, hallmarked by a highly polymorphic alpha chain called HLA-A, HLA-B and HLA-C, that all present epitopes to cytolytic T cells. 
For MHC-II, both the alpha and the beta chains contribute to the peptide binding groove. There are three classical forms of MHC-II as well, called HLA-DR, HLA-DQ and HLA-DP, that all present epitopes to helper T cells.
Each MHC complex can present a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
epitopes they bind (\cite{lund2004definition}).
Moreover, the HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA allele is called 
an MHC haplotype and presents a different subset of peptides (\cite{marsh2010nomenclature}).

@@

Humans mostly express two haplotypes per MHC form, one from the parental and one from the maternal chromosome, and therefore an individual's immune system detects 
only a fraction of all possible peptide fragments. 
However, at the population level, the coverage of pathogenic peptides that are detected 
is very high, because of the highly polymorphic MHC genes.
It is therefor believed that MHC polymorphism improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation at the individual level, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes present in the population  (\cite{sommer2005importance}).

@@

Many studies are aimed at identifying the repertoire of epitopes that are presented in any MHC haplotype and determining which epitopes will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development of prediction algorithms 
that allow for very reliable \emph{in silico} predictions 
of the binding affinities of peptides
(\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).
For example, \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to MHC,
86\% were experimentally confirmed to do so.

@@

Using these prediction algorithms, 
we recently predicted that peptides derived 
from transmembrane helices (TMHs) 
will be more frequently presented by MHC-I 
than expected based on their abundance (\cite{bianchi2017}). 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs. 
This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity 
to bind than more hydrophilic peptides.

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted with high accuracy from a protein sequence 
by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}),
for example, \cite{jones2007improving} found that,
from 184 transmembrane proteins (TMPs) with known topology, 
80\% of the TMH predictions replicated this finding.

TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH (\cite{ahram2006estimation}).
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble (i.e., extracellular or cytoplasmic) protein regions is unknown. 
We hypothesized that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutations. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMHs 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

This study had two objectives. 
First, we aimed to generalize our findings by predicting the presentation of peptides 
from different kingdoms of life and for both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, 
and for most haplotypes of both MHC-I and II. 
We confirmed the presentation of TMH-derived peptides 
by re-analysis of peptide elution studies. 
Second, we tested our hypothesis that TMHs 
are more evolutionary conserved than soluble protein regions. 
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell branch of the adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

To predict how frequently epitopes overlapping with TMHs are presented,
a similar analysis strategy was applied as described in \cite{bianchi2017} 
for several haplotypes of both MHC-I and MHC-II, 
and for a human, viral and bacterial proteome.
To summarize, for each proteome, 
all possible 9-mers (for MHC-I) or 14-mers (MHC-II) were derived. 
For each of these peptides, we determined if it overlapped with a predicted 
TMH and if it was predicted to bind to each haplotype.

For MHC-I, 9-mers were used, as this is the length most frequently presented in MHC-I and was used in our earlier study (\cite{bianchi2017}). 
For MHC-II, 14-mers were used, 
as these are the most frequently occurring epitope length (\cite{bergseng2015different}).
A human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis}, UniProt ID UP000001584) 
reference proteome was used. 
\verb;TMHMM; (\cite{krogh2001predicting}) was used to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain MHC haplotype,
 \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; (\cite{shao2020high}) for MHC-II was used.
The 13 MHC-I haplotypes used in this study are the same as 
used in the previous study (\cite{bianchi2017}).
For MHC-II, haplotypes were selected with a phenotypic frequency 
of at least 14\% in the human population (\cite{greenbaum2011functional}),
resulting in 21 MHC-II haplotypes.

\geert{wat text gewist want dit is de methods sectie}
In previous work, it was found that the over-presentation of TMH-derived peptides can be
explained from the hydrophobicity of the MHC-I binding cleft (\cite{bianchi2017}).
Here, a similar analysis was applied, by correlating the percentage of predicted TMH-derived epitopes versus the mean hydrophobicity of all 
peptides.

This study differs in one important aspect from our previous work (\cite{bianchi2017}). 
The definition of a binder differs from \cite{bianchi2017}:
in the current study, a peptide is called a binder if, for a certain haplotype, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
whereas the previous study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the results are similar, a side by side comparison was performed shown in the supplementary materials. \geert{refereer hier naar supplementary figuur}

@@

To obtain experimental evidence that epitopes derived from TMHs are presented in MHC, peptide elution studies for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}) were reanalyzed.
For each of the detected epitopes,
its possible location(s) in 
a human reference proteome,
with UniProt ID UP000005640\_9606, was mapped.
For the epitopes that were present in the proteome exactly once,
the topology of the proteins in which these epitopes were located was predicted using both \verb;TMHMM; (\cite{krogh2001predicting}) and \verb;PureseqTM; (\cite{wang2019efficient}). 
From this topology, we determined if the epitope
overlapped with a TMH.

@@

To determine the evolutionary conservation of TMHs,
human single nucleotide polymorphisms (SNPs) were first collected
that resulted in a single amino acid substitution,
and we then determined if this substitution occurred within a predicted TMH or not.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: the \emph{dbSNP} (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases \emph{gene} (for gene names, \cite{brown2015gene})
and \emph{protein} (for proteins sequences, \cite{sayers2010database}).

@@

The first query was a call to the \emph{gene} database for the 
term 'membrane protein' (in all fields) 
for the organism \emph{Homo sapiens}.
This resulted in 1,077 gene IDs (on December 2020).
% 1077 gene IDs is correct for December 2020.
% At 2021-03-01, one will get 1130 gene IDs.
% Also, one of the gene IDs that was valid back then,
% has been obsoleted.
The next query was a call to the \emph{gene} database 
to obtain the gene names from the gene IDs.
Per gene name, the \emph{dbSNP} NCBI database was queried for 
variations associated with the gene name. 
As the NCBI API constrains its users to three calls per second
(to assure fair use), we had to limit the extent of our analysis.

@@

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that these ID numbers are unrelated to SNP positions, the relative positions of all analyzed SNPs in a protein were determined. This analysis showed no positional bias of the SNPs, as shown in \geert{dit is supplement toch?} supplementary figure \ref{fig:snp_rel_pos}.

Per SNP, the \emph{protein} NCBI database was queried for the
protein sequence. For each protein sequence, the protein topology was determined using \verb;PureseqTM;. Using these predicted protein topologies, the SNPs were scored to be located within or outside TMHs.

@@

Figure \ref{fig:bbbq_1_smart_results} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I haplotype, it shows the percentage of binders that overlap with a TMH 
with at least one residue.
The horizontal line shows the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as epitopes derived from soluble regions.
For 11 out of 13 MHC-I haplotypes, TMH-derived epitopes are predicted to be presented more often 
than the null expectation, for a human and bacterial proteome.
For the viral proteome, 12 out of 13 haplotypes present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the different haplotypes 
is similar for the probed proteomes, 
which strengthens our previous conclusion (\cite{bianchi2017}) 
that the hydrophobicity of the MHC-binding groove 
is the main factor responsible for the predicted over-presentation 
of TMH-derived peptides.  \frans{zoals eerder HLA-C mist en ook HLA-DP voor de nu gekozen haplotypes mis ik de rationale waarom we deze hebben genomen? volgens mij had ik een artikel gestuurd met de meest voorkomende HLA-DP haplotypes. De rationale voor HLA-C is dat de expressie hier naar verhousding heel laag van is en dat dus een minder explecitie rol speelt in antigen presentatie} \geert{in methods section geeft Richel al rationale voor de keuze van de haplotypes, waarin hij schrijft dat dit gebasseerd is op onze eerdere studie (voor MHC-I) en de frequentie van de haplotypes (voor MHC-II). Dit lijkt mij voldoende.}

@@

  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc1.png}
    \label{fig:hydrophobicity_1}
  \end{subfigure}  
  \hfill
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc2.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \caption{ \textbf{Over-presentation of TMH-derived epitopes on most MHC-I and -II haplotypes} \geert{een figuur heeft altijd een titel met de conclusie. De legenda zelf is feitelijke opsomming en biedt geen (of weinig) interpretatie}
    \textbf{(A)} 
    The percentage of epitopes for MHC-I and -II haplotypes that are predicted to 
    overlap with TMHs for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The red lines indicate the percentages as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \textbf{(B-C)}
    Correlation between the percentages of predicted TMH-derived epitopes
    and the hydrophobicity score of all predicted epitopes for MHC-I \textbf{(B)}
    and MHC-II haplotypes \textbf{(C)}. Red curve: linear regression analysis. Labels are shorthand for the HLA haplotypes,
    see Table \ref{tab:haplotype_abbreviations} for the names. \frans{wat is de reden dat nr 15 en 8 er buiten vallen? de rest is vrijwel linear)
  }}
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II haplotypes, 
for a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{M. tuberculosis}.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the exact TMH and epitope counts.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The over-presentation of TMH-derived peptides is caused by the hydrophobicity of the MHC peptide binding groove}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For MHC-I, we previously showed that the over-presentation of TMH-derived peptides is caused by the hydrophobicity of the peptide binding grooves \cite{bianchi2017}. Figures \label{fig:hydrophobicity_1} and \label{fig:hydrophobicity_2} \frans{de verwijzing werkt denk ik niet}
show the extent of over-presentation
of TMH-derived epitopes as a function of the hydrophobicity preference score for the different haplotypes.
An assumed linear correlation explains 88\% of the variability in MHC-I.
For MHC-II, 62\% of the variability is explained by hydrophobicity. \frans{dit is dus omdat 2 haplotypes buiten de boot vallen} This strengthens our previous finding \cite{bianchi2017} and indicates that TMH-derived peptides are over-presented because the peptide binding grooves of most MHC-I and -II haplotypes are relatively hydrophobic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experimental validation of presentation of TMH-derived peptides}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{bbbq_article_issue_157/results.png}
  \caption{ \textbf{MHC-I and -II presentation of TMH-derived peptides in peptide elution studies.}
    TMH-derived epitopes are presented \emph{in vivo},
    for both MHC-I and MHC-II, for two topology prediction
    programs. \geert{legenda is veel te summier. Wat is precies geplot? ook studies van de data citeren.}
  }
  \label{fig:elution}
\end{figure}

To obtain experimental confirmation that peptides stemming from TMHs 
are presented in MHC-I and MHC-II,
two peptide elution studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} for MHC-I and MHC-II respectively) 
were reanalyzed.
Figure \ref{fig:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I and MHC-II elution 
studies, \geert{is dit van totale peptides, of van humane peptides? Uiteraard van humane peptides gebruiken, want de rest heb je niet gemapped. Dat moet dan wel hier gezegd worden. Daarnaast: zijn de MHC-I en -II haplotypes bekend in deze studies?}
for the two topology prediction tools \verb;TMHMM; (\cite{krogh2001predicting}) and \verb;PureseqTM; (\cite{wang2019efficient}). 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.
See the supplementary materials (table \ref{tab:elution}) for the exact values.

@@

Together, from this study, two important conclusions can be drawn. 
First, the MHC over-presentation of TMHs is likely a general feature 
and predicted to occur for most haplotypes of both MHC-I and -II 
and for humans as well as bacterial and viral pathogens. 
Second, TMHs are genuinely more evolutionary conserved than soluble protein motifs, 
at least in the human proteome.
